RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-05 manuscripts/manuscript.Rmd
Tightening up various places in package description.

The `RNeXML` package is written entirely in R and available under a
CC0 public domain waiver. The current development version can be found
on Github at [https://github.com/ropensci/RNeXML](), and the stable
version can be installed from the CRAN repository:

```r
install.packages("RNeXML", dependencies=TRUE)
```

This will also install its dependencies. Once installed, R scripts
using the library only need to load it:

```{r echo=FALSE}
library(RNeXML)
```

@@

Conceptually, a NeXML document has the following components: (1)
phylogeny data, (2) character or trait data in matrix form, (3)
operational taxonomic units (OTUs), and (4) metadata. `RNeXML` defines
the `nexml` object type, which is a representation of the NeXML
document contents. It therefore holds phylogenetic trees as well as
character or trait matrices, and all metadata. This is similar in
design to the classes in the `phylobase` package, which contain trees
and traits but hold only limited metadata [@phylobase], but different
from R object types such as `ape::phylo` (the `phylo` class in the
`ape` package [@Paradis_2004]), which represents trees alone.

When reading and writing NeXML documents, `RNeXML` aims to map their
components to and from, respectively, their most natural
representations in R. Hence, the types of objects accepted or returned
by the package's methods are the `phylo` and `multiPhylo` objects from
the `ape` package [@Paradis_2004] for phylogenies, and R's native
`data.frame` list structure for character matrices.

@@

The method `nexml_read()` reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type `nexml` [^2]:

@@

To interface conveniently with the many already existing R packages for 
phylogenetics, which conventionally accept phylogenies as `ape::phylo`
objects, the method `get_trees_list()` returns a `ape::multiPhylo`
object, which can be treated as a list of `ape::phylo` objects:

@@

The `get_trees_list()` method is designed for use in scripts, for
which a consistent and predictable return type (namely a `list` of
`ape::multiPhylo` objects, reflecting the nested structure of NeXML)
is desirable. `RNeXML` also contains a method `get_trees()`, which
is designed for interactive use and returns the R object most
intuitive given the arrangement of phylogeny data in the source NeXML
document. NeXML inherits from NEXUS the ability to contain multiple
phylogenies, and to group sets of phylogenies into "blocks". The
`get_trees()` method will return an `ape::phylo` object if the NeXML
document contains a single phylogeny, an `ape::multiPhylo` object if
it contains multiple phylogenies arranged in a single `trees` block,
and a list of `ape::multiPhylo` objects if it contains multiple
`trees` blocks. This design spares an interactive user from having to
type unnecessary data conversion commands, at the expense of needing
to know which return type to expect (or verifying the type via
introspection commands such as `class()`).

If the location parameter with which the `nexml_read()` method is
invoked is recognized as a URL, the method will automatically download the
document to the local working directory and read it from there. This
gives convenient and rapid access to phylogenetic data published in
NeXML format on the web, such as the content of the
phylogenetic data repository TreeBASE [@Piel_2002]. For
example, the following plots a tree in TreeBASE:

```{r eval=FALSE}
tb_nex <- nexml_read(
"https://raw.github.com/TreeBASE/supertreebase/master/data/treebase/S100.xml")
tb_phy <- get_trees_list(tb_nex)
```

The method `get_characters()` obtains character data matrics from a
`nexml` object, and returns them as a standard `data.frame` R object
with columns as characters and rows as taxa:

@@

A NeXML data matrix can be of molecular (for molecular sequence
alignments), discrete (for most morphological character data), or
continuous type (for many trait data).  Data matrices can be, and
matrices of different data types are required to be separated into
different "blocks", to allow strict validation of data types. Since
the `data.frame` data structure in R has no such constraints, the
`get_characters()` method will combine such blocks as separate columns
into a single `data.frame` object, provided they correspond to the
same taxa. Otherwise, a list of `data.frame`s is returned, with list
elements corresponding to characters blocks. Similar to the methods
for obtaining trees, there is also a method `get_characters_list()`,
which always returns a list of `data.frame`s, one for each character
block.

@@

The first argument to `nexml_write()` is an object of type `nexml`, or any
object that can be coerced to it, such as in the above example an
`ape::phylo` phylogeny. Alternatively, passing a `multiPhylo` object
would write a list of phylogenies to the file.

@@

Instead of packaging the various components for a NeXML file at the
time of writing the file, `RNeXML` also allows users to create and
iteratively populate an in-memory `nexml` object. The methods to do
this are `add_characters()`, `add_trees()`, and `add_meta()`, for
adding characters, trees, and metadata, respectively. Each of these
functions will automatically create a new nexml object if not supplied
with an existing one as the last (optional) argument.

@@

The data with which a `nexml` object is populated need not share the
same OTUs.  `RNeXML` automatically adds new, separate OTU blocks into
the NeXML file for each data matrix and tree that uses a set of OTUs
that differs from others.

Other than storage size, there is no limit to the number of
phylogenies and character matrices that can be included in a
single NeXML file. This allows, for example, to capture samples from a
posterior probability distribution of inferred or simulated phylogenies or
character states in one NeXML file.

@@

NeXML allows attaching metadata ("_annotations_") to any data element,
and even to metadata themselves. Whether at the level of the document
as a whole or the an individual data matrix or phylogeny, metadata can
provide bibliographic and provenance information, for example about
the study as part of which the phylogeny was generated or applied,
which data matrix and which methods were used to generate it. Metadata
can also be attached to very specific elements of the data, such as
specific traits, individual OTUs, nodes, or even edges of the
phylogeny.

NeXML uses the "Resource Description Framework in Annotations" (RDFa),
as described in @Vos_2012 or @W3C_2014, to encode metadata
annotations. This standard provides a precisely specified,
machine-readable format that maintains backwards compatibility with
compliant NeXML parsers (and thus `RNeXML`), because the capacity of a
tool to _parse_ annotations is not predicated on _understanding_
the meaning of annotations it has not seen before.

The burden typically involved with documenting and sharing good
metadata alongside the data remains a significant barrier to increased
sharing and reuse of scientific research [@Tenopir_2011]. To promote
sharing of well-documented phylogenetic data, `RNeXML` aims to
alleviate the burden to recording useful and machine-readable metadata
at several levels.

First, when writing a NeXML file the package adds certain basic
metadata automatically if it has not yet been provided, using default
values consistent with recommended best practices
[@Cranston_2014]. Currently, this includes naming the software
generating the NeXML, a time-stamp of when a tree was produced, and an
open data license.

Second, `RNeXML` provides a simple method, called
`add_basic_metadata()`, to set metadata attributes commonly
recommended to be included with publicly archived or shared data
[@Cranston_2014]. The currently accepted parameters include `title`,
`description`, `creator`, `pubdate`, `rights`, `publisher`, and
`citation`. Behind the scenes the method automatically anchors these
attributes in common vocabularies (such as Dublin Core).

Third, `RNeXML` includes methods to assist with common cases of adding
metadata to specific elements of the NeXML data that increase the
data's utility for reuse. For example, misspellings and inconsistent
taxonomic names are a frequent source of frustration especially when
recombining phylogenetic data from different sources. These obstacles
can be mitigated by annotating the taxonomic names with identifiers,
which also allows being explicit about the identifier authority being
used. `RNeXML` includes the `taxize_nexml()` method, which uses the R
package `taxize` [@Chamberlain_2013] to match each taxon label against
the NCBI database, and, provided a unique match is found, annotates
the respective OTU with the matching NCBI identifier,.

@@

The RDFa standard adopted by NeXML (see @Vos_2012) for all metadata
annotations defines the structure and syntax for such annotations, but
does not restrict the content. Specifically, every metadata annotation
consists of a metadata property and a value of the property; the
bearer of the property is determined by where the annotation appears
in the NeXML document (technically, the containing element in the XML
document model). Technically, this structure corresponds directly to
Resource Description Framework (RDF) subject-predicate-object triples.
It therefore allows for arbitrary custom extensions to meet ongoing
scientific needs without requiring modifications (or "extensions") of
the syntax, provided the metadata annotations are cast as RDF triples.

The `RNeXML` methods described above for built-in metadata annotations
perform this cast automatically behind the scenes, including mapping
metadata attributes to terms in requisite common vocabularies (such as
Dublin Core for "title", "creator", etc.) or ontologies. Giving
metadata properties (and sometimes values) as vocabulary or ontology
terms rather than simple text strings is crucial for allowing machines
to not only parse but also interpret and potentially reason over their
semantics. To achieve this benefit for custom metadata extensions, the
user necessarily needs to handle certain technical details from which
the `RNeXML` API shields her otherwise, in particular the vocabularies
or ontologies of metadata terms in the form of globally unique
identifiers (normally HTTP URIs). To be consistent with XML
terminology for more technical users, `RNeXML` calls these
_namespaces_, and their abbreviations _prefixes_. For example, the
namespace for the Dublin Core Metadata Terms vocabulary, commonly
abbreviated with "dc", is "http://purl.org/dc/elements/1.1/". Using
its abbreviation, a metadata property "dc:title" expands to the
identifier "http://purl.org/dc/elements/1.1/title", the identifier for
the term "title" in said vocabulary. When resolved as a HTTP URI, the
identifier resolves to a human and machine-readable
(depending on how the identifier is resolved) definition of precisely
what the term `title` in Dublin Core means. In contrast, just using
the text string "title" could also mean the title of a person, a legal title, the verb title, etc.  Although URI identifiers of metadata
vocabularies and terms are not required to resolve, for machines to
derive the maximum benefit from them, they should resolve to RDF
statements defining their semantics.

`RNeXML` includes methods to obtain and manipulate metadata
properties, values, identifiers, and namespaces. The
`get_namespaces()` method accepts a `nexml` object and returns a list
with element names being the namespace abbreviations and element
values being the corresponding identifiers (typically HTTP URIs) for
all annotations attached to the object or one of the contained data elements:

@@

The `get_metadata()` method returns, as named list, the metadata
annotations for a given `nexml` object at a given level, with the
whole NeXML document being the default:

```{r}
meta <- get_metadata(birds) 
```

The `level` parameter can be used to extract the metadata objects for
the data at specific levels (such as `"otu"` in the following
example), or by specifying `"all"` to extract all metadata objects:

```{r} 
otu_meta <- get_metadata(birds, level="otu")
```

However, because the returned list does not include the data elements
to which the metadata are attached, using the method in this way is
only useful for identifying which metadata are present and their
values. The recommended approach for accessing the metadata attached
to specific data elements is described below. <!-- or is that in the vignette now? -->

A few select metadata properties for which support is built-in have
dedicated accessor methods, for example the citation:

```{r get_citation}
get_citation(birds)
```

The `meta()` method creates a new metadata object from a poperty name
and content (value). For example, the following creates a modification
date metadata object, using the modificationDate property in the PRISM
vocabulary:

@@

The `add_meta()` function adds metadata objects as annotations to a
`nexml` object at a specified level. The default level is the NeXML document as a whole:

@@

If the prefix used by the metadata property is not among the built-in
ones (which can be obtained using the `get_namespaces()` as described
above), it has to be provided along with its URI as the `namespaces`
parameter. For example, the following uses the SKOS vocabulary (which
currently is not built-in) to add a note to the trees in the `nexml`
object:

@@

Alternatively, additional namespaces can also be added in batch using
the `add_namespaces()` method, which makes a list of namespaces known
to a given `nexml` object (or created a new `nexml` object, similar to
other `add_*` methods).

`RNeXML`, by virtue of subsetting the S4 `nexml` object, also offers
fine control of the level at which a `meta` element is added, for
which the package vignette contains examples.

Because NeXML expresses all metadata using the RDF standard, and
stores them compliant with RDFa, they can be extracted as an RDF graph
using off-the-shelf tools, and the RDF graph obtained in this way can
be queried in SPARQL (see @W3C_2014 or @Hartig_2012), the query
language for RDF graphs, using general purpose triple stores and
SPARQL query engines. The `RNeXML` vignette illustrates this with
examples, and the package also comes with a demonstration that can be
run from R using the following command: `demo("sparql", "RNeXML")`).

@@

No data standard can anticipate all the uses and extensions
researchers may one day need. As the various "flavors" of the NEXUS
and other data formats show, this often results in extensions to
widely used exchange formats that almost invariably break
interoperability between such extensions and tools that were designed
to read one but not another. NeXML was designed to prevent this by
means of a highly flexible metadata system, while not sacrificing 
strict validation of syntax and structure.

Here we illustrate how NeXML's metadata system, and `RNeXML`'s API to
it, can be used to record and share a type of phylogenetic data not
taken into account when NeXML was designed. Specifically, we use the
example of stochastic character maps [@Huelsenbeck_2003], which assign
certain parts (corresponding to time) of each branch in a
time-calibrated phylogeny to a particular "state" (typically of a
morphological characteristic). The current de-facto format for sharing
stochastic character maps is a (non-interoperable) modification of the
standard Newick tree format created by a software tool widely used for
creating such maps, `simmap` [@Bollback_2006].

To express stochastic character map data as a custom metadata
extension, we create metadata annotations for the `edge` elements in
the phylogeny topology that relate parts of the annotated edge (branch)
to different states. These annotations need to describe the character state
being assigned, and the duration (in terms of branch length) that the edge 
spends in that state. The `NeXML` standard already defines `state`
elements for discrete character traits, which, as most elements in
NeXML, also have unique identifies (attribute `id`) by which they can
be referenced in other parts of the document. In the following
proposal, we call each series of character state assignments for an edge a
_reconstruction_, and for each state assignment comprising the
reconstruction, we provide the length, the identifier of the character
state, and an ordinal property to record the chronological order of
the state assignment (the identifier values in the proposal, "cr1",
"s1", and "s2", are -- fictitious -- identifiers that, in contrast to
global identifiers, are unique only within a given NeXML document):

@@

As we illustrate here, even if the URI does not resolve to a fully
machine-readable definition of the terms and their semantics, it can
still be used to provide at least a human-readable informal definition
of the terms, and where to find such documentation _is_
machine-readable in the resulting NeXML document. Aside from the fact
that the above representation does not modify the NeXML format, and
should thus be readable -- even if not necessarily interpretable as a
stochastic character mapping -- by any compliant NeXML parser, the
direct and machine-readable link to metadata documentation contrasts
with metadata or format extensions that are undocumented, or which do
not directly link to documentation.

Although writing out metadata annotations as in the above proposal may
seem tedious, by virtue of `RNeXML`'s API they can be expressed as a
series of R commands, and it is therefore straightforward to define
functions that encapsulate the details behind a parameter
signature. To demonstrate this, we implemented two functions in
`RNeXML`, `simmap_to_nexml()` and `nexml_to_simmap()`, that map
bidirectionally between NeXML files with `simmap` metadata according
to the above proposal, and the `simmap` data representation in R
devised by @Revell_2012 in the R package `phytools`. The latter uses
an extension of the `ape::phylo` class. As shown in more detail in the
`RNeXML` package vignette, this allows generating NeXML files
containing `simmap` data, reading those files into R, converting them
into the in-memory representation understood by `phytools`, and using
the various functions from `phytools` designed for `simmap` objects
[@Revell_2012], such as `plotSimmap()` for plotting.

@@

One of `RNeXML`'s aims is to promote the archival of well-documented
phylogenetic data in scientific data repositories, in the form of
NeXML files. To this end, the method `nexml_publish()` provides an API
directly from within R that allows data archival to become a step programmed
into data management scripts, rather than a separate manually executed
act. Initially, the method supports the data repository Figshare (http://figshare.com):

>>>>

The `RNeXML` package is written entirely in R and available under a
Creative Commons Zero public domain waiver. The current development version can be found
on Github at [https://github.com/ropensci/RNeXML](), and the stable
version can be installed from the CRAN repository.

```{r echo=FALSE}
install.packages("RNeXML", dependencies=TRUE)
library(RNeXML)
```

@@

Conceptually, a NeXML document has the following components: (1)
phylogeny topology data, (2) character or trait data in matrix form, (3)
operational taxonomic units (OTUs), and (4) metadata. To represent the
contents of a NeXML document (currently in memory), `RNeXML` defines
the `nexml` object type. This type therefore holds phylogenetic trees
as well as character or trait matrices, and all metadata, which is
similar to the phylogenetic data object types defined in the
`phylobase` package [@phylobase], but contrasts with the more widely
used ones defined in the
`ape` package [@Paradis_2004], which represents trees alone.

When reading and writing NeXML documents, `RNeXML` aims to map their
components to and from, respectively, their most widely used
representations in R. As a result, the types of objects accepted or returned
by the package's methods are the `phylo` and `multiPhylo` objects from
the `ape` package [@Paradis_2004] for phylogenies, and R's native
`data.frame` list structure for data matrices.

@@

The method `nexml_read()` reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type `nexml`:

@@

The method `get_trees_list()` can be used to extract the phylogenies
as an `ape::multiPhylo`
object, which can be treated as a list of `ape::phylo` objects:

@@

The `get_trees_list()` method is designed for use in scripts, providing
a consistent and predictable return type regardless of the number of
phylogenies a NeXML document contains. For greater convenience in
interactive use, the method `get_trees()` returns the R object most
intuitive given the arrangement of phylogeny data in the source NeXML
document. For example, the method returns an `ape::phylo` object if the NeXML
document contains a single phylogeny, an `ape::multiPhylo` object if
it contains multiple phylogenies arranged in a single `trees` block,
and a list of `ape::multiPhylo` objects if it contains multiple
`trees` blocks (the capability for which NeXML inherits from NEXUS).

If the location parameter with which the `nexml_read()` method is
invoked is recognized as a URL, the method will automatically download the
document to the local working directory and read it from there. This
gives convenient and rapid access to phylogenetic data published in
NeXML format on the web, such as the content of the
phylogenetic data repository TreeBASE [@Piel_2002]. For
example, the following plots a tree in TreeBASE (using ape's plot function):

```{r eval=FALSE fig.show="hide"}
tb_nex <- nexml_read(
"https://raw.github.com/TreeBASE/supertreebase/master/data/treebase/S100.xml")
tb_phy <- get_trees_list(tb_nex)
plot(tb_phy[[1]]) 
```

The method `get_characters()` obtains character data matrices from a
`nexml` object, and returns them as a standard `data.frame` R object
with columns as characters and rows as taxa:

@@

A NeXML data matrix can be of molecular (for molecular sequence
alignments), discrete (for most morphological character data), or
continuous type (for many trait data). To enable strict validation of data types NeXML allows, and if their
data types differ requires multiple data matrices to be separated into
different "blocks". Since
the `data.frame` data structure in R has no such constraints, the
`get_characters()` method combines such blocks as separate columns
into a single `data.frame` object, provided they correspond to the
same taxa. Otherwise, a list of `data.frame`s is returned, with list
elements corresponding to characters blocks. Similar to the methods
for obtaining trees, there is also a method `get_characters_list()`,
which always returns a list of `data.frame`s, one for each character
block.

@@

The first argument to `nexml_write()` is either an object of type `nexml`, or any
object that can be coerced to it, such as in the above example an
`ape::phylo` phylogeny. Alternatively, passing a `multiPhylo` object
would write a list of phylogenies to the file.

@@

Instead of packaging the various components for a NeXML file at the
time of writing the file, `RNeXML` also allows users to create and
iteratively populate in-memory `nexml` objects. The methods to do
this are `add_characters()`, `add_trees()`, and `add_meta()`, for
adding characters, trees, and metadata, respectively. Each of these
functions will automatically create a new nexml object if not supplied
with an existing one as the last (optional) argument.

@@

The data with which a `nexml` object is populated need not share the
same OTUs.  `RNeXML` automatically adds new, separate OTU blocks into
the NeXML file for each data matrix and tree that uses a different set of OTUs.

Other than storage size, there is no limit to the number of
phylogenies and character matrices that can be included in a
single NeXML document. This allows, for example, to capture samples from a
posterior probability distribution of inferred or simulated phylogenies and
character states in a single NeXML file.

@@

NeXML allows attaching ("_annotating_") metadata to any data element,
and even to metadata themselves. Whether at the level of the document
as a whole or an individual data matrix or phylogeny, metadata can
provide bibliographic and provenance information, for example about
the study as part of which the phylogeny was generated or applied,
which data matrix and which methods were used to generate it. Metadata
can also be attached to very specific elements of the data, such as
specific traits, individual OTUs, nodes, or even edges of the
phylogeny.

As described in @Vos_2012,  to encode metadata annotations NeXML uses
the "Resource Description Framework in Annotations" (RDFa)
[@W3C_2014]. This standard provides for a strict machine-readable
format yet enables future backwards compatibility with
compliant NeXML parsers (and thus `RNeXML`), because the capacity of a
tool to _parse_ annotations is not predicated on _understanding_
the meaning of annotations it has not seen before.

To lower the barriers to sharing well-documented phylogenetic data,
`RNeXML` aims to make recording useful and machine-readable metadata
easier at several levels.

First, when writing a NeXML file the package adds certain basic
metadata automatically if they are absent, using default
values consistent with recommended best practices
[@Cranston_2014]. Currently, this includes naming the software
generating the NeXML, a time-stamp of when a tree was produced, and an
open data license.

Second, `RNeXML` provides a simple method, called
`add_basic_metadata()`, to set metadata attributes commonly
recommended for inclusion with data to be publicly archived or shared
[@Cranston_2014]. The currently accepted parameters include `title`,
`description`, `creator`, `pubdate`, `rights`, `publisher`, and
`citation`. Behind the scenes the method automatically anchors these
attributes in common vocabularies (such as Dublin Core).

Third, `RNeXML` integrates with the R package `taxize`
[@Chamberlain_2013] to mitigate one of the most common obstacles to
reuse of phylogenetic data, namely the misspellings and inconsistent
taxonomic naming with which OTU labels are often fraught. The
`taxize_nexml()` method in `RNeXML` uses `taxize` to match OTU labels
against the NCBI database, and, where a unique match is found, it
annotates the respective OTU with the matching NCBI identifier.

@@

The RDFa standard adopted by NeXML (see @Vos_2012) allows for
arbitrary custom extensions to meet ongoing scientific needs without
requiring modifications (or "extensions") of the syntax. Specifically,
every metadata annotation consists of a metadata property and a value
of the property; the bearer of the property is determined by where the
annotation appears in the NeXML document (technically, the containing
element in the XML document model). This structure corresponds
directly to Resource Description Framework (RDF)
subject-predicate-object triples, and therefore the only requirement
for custom metadata is that they be cast as RDF triples.

The `RNeXML` interface described above for built-in metadata annotations
perform this cast automatically behind the scenes, including mapping
metadata attributes to terms in requisite common vocabularies (such as
Dublin Core for "title", "creator", etc.) or ontologies. Giving
metadata properties (and where applicable, values) as vocabulary or ontology
terms rather than simple text strings is crucial for allowing machines
to not only parse but also interpret and potentially reason over their
semantics. To achieve this benefit for custom metadata extensions, the
user necessarily needs to handle certain technical details from which
the `RNeXML` API shields her otherwise, in particular the globally unique
identifiers (normally HTTP URIs) of metadata terms and vocabularies. To be consistent with XML
terminology, `RNeXML` calls vocabulary URIs
_namespaces_, and their abbreviations _prefixes_. For example, the
namespace for the Dublin Core Metadata Terms vocabulary is "http://purl.org/dc/elements/1.1/". Using
its common abbreviation "dc", a metadata property "dc:title" expands to the
identifier "http://purl.org/dc/elements/1.1/title". This URI resolves
to a human and machine-readable (depending on access) definition of precisely
what the term `title` in Dublin Core means. In contrast, just using
the text string "title" could also mean the title of a person, a legal
title, the verb title, etc.  URI identifiers of metadata vocabularies
and terms are not mandated to resolve, but if machines are to derive
the maximum benefit from them, they should resolve to a definition of
their semantics in RDF.

`RNeXML` includes methods to obtain and manipulate metadata
properties, values, identifiers, and namespaces. The
`get_namespaces()` method accepts a `nexml` object and returns a named list
of namespace prefixes and their corresponding identifiers known to the object:

@@

The `get_metadata()` method returns, as a named list, the metadata
annotations for a given `nexml` object at a given level, with the
whole NeXML document being the default level (`"all"` extracts all metadata objects):

```{r}
meta <- get_metadata(birds) 
otu_meta <- get_metadata(birds, level="otu")
```

The returned list does not include the data elements to which the
metadata are attached. Therefore, a different approach, documented in
the package vignette, is recommended for accessing the metadata
attached to data elements.

The `meta()` method creates a new metadata object from a poperty name
and content (value). For example, the following creates a modification
date metadata object, using a property in the PRISM vocabulary:

@@

The `add_meta()` function adds metadata objects as annotations to a
`nexml` object at a specified level, with the default level being the
NeXML document as a whole:

@@

If the prefix used by the metadata property is not among the built-in
ones (which can be obtained using `get_namespaces()`), it has to be provided along with its URI as the `namespaces`
parameter. For example, the following uses the SKOS vocabulary (which
currently is not built-in) to add a note to the trees in the `nexml`
object:

@@

Alternatively, additional namespaces can also be added in batch using
the `add_namespaces()` method.

By virtue of subsetting the S4 `nexml` object, `RNeXML` also offers
fine control of where a `meta` element is added, for which the package
vignette contains examples.

Because NeXML expresses all metadata using the RDF standard, and
stores them compliant with RDFa, they can be extracted as an RDF
graph, queried, analyzed, and mashed up with other RDF data, local or
on the web, using a wealth of off-the-shelf tools for working with RDF
(see @W3C_2014 or @Hartig_2012). Examples for these possibilities are
included in the `RNeXML` vignette, and the package also comes with a
demonstration that can be run from R using the following command:
`demo("sparql", "RNeXML")`).

@@

NeXML was designed to prevent the need for future non-interoperable
"flavors" of the standard in response to new research directions. Its
solution to this inevitable problem is a highly flexible metadata
system without sacrificing strict validation of syntax and structure.

Here we illustrate how `RNeXML`'s interface to NeXML's metadata system
can be used to record and share a type of phylogenetic data not taken
into account when NeXML was designed, in this case stochastic
character maps [@Huelsenbeck_2003]. Such data assign certain parts
(corresponding to time) of each branch in a time-calibrated phylogeny
to a particular "state" (typically of a morphological
characteristic). The current de-facto format for sharing stochastic
character maps, created by `simmap` [@Bollback_2006], a widely used tool
for creating such maps, is a non-interoperable modification of the standard
Newick tree format.

To express these data as a custom metadata
extension, we create metadata annotations for the `edge` elements in
the phylogeny topology. The metadata need to describe the character state
being assigned, and the duration (in terms of branch length) that the edge 
spends in that state. The `NeXML` standard already defines `state`
elements for discrete character traits, which, as most elements in
NeXML, also have unique identifiers by which they can
be referenced. In the following
proposal, we term each series of character state assignments for an edge a
_reconstruction_, and for each state assignment comprising a
reconstruction, we provide the length, the identifier of the character
state, and an ordinal property to record the chronological order of
the assignment (the identifier values in the proposal, "cr1",
"s1", and "s2", are identifiers local to a  -- fictitious -- NeXML document):

@@

Here the URI does not resolve to a fully
machine-readable definition of the terms and their semantics, but it can
nonetheless be used to provide at least a human-readable informal definition
of the terms.

Because the above representation does not modify the NeXML format, it
should be readable -- even if not necessarily interpretable as a
stochastic character mapping -- by any compliant NeXML parser. In
addition, it includes a direct and machine-readable link to metadata
documentation.

Writing out metadata annotations as in the above proposal may seem
tedious. However, by virtue of `RNeXML`'s API it is straightforward to
define functions that encapsulate these details. As a demonstration,
`RNeXML` comes with functions `simmap_to_nexml()` and
`nexml_to_simmap()` that map bidirectionally between NeXML files with
`simmap` metadata according to the above proposal, and the extension
of the `ape::phylo` class devised by @Revell_2012 in the R package
`phytools`. The `RNeXML` package vignette illustrates how this allows
seamlessly generating and reading NeXML files containing `simmap`
data, and using the various functions from `phytools` designed for
`simmap` objects [@Revell_2012].

@@

One of `RNeXML`'s aims is to promote the archival of well-documented
phylogenetic data in scientific data repositories, in the form of
NeXML files. To this end, the method `nexml_publish()` provides an API
directly from within R that allows data archival to become a step programmed
into data management scripts. Initially, the method supports the data repository Figshare (http://figshare.com):
